\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-11-30 chapters/abstract_nl.tex
Preface, ironed out overhangs

Naast onafhankelijkheid van oogbewegingen, verbetert dit werk ook algemene
ERP decodering door de structuur van lineaire en multilineaire technieken te verfijnen.
Deze methoden verhogen de nauwkeurigheid in verschillende BCI-toepassingen,
vooral wanneer kalibratiegegevens beperkt zijn.
Gestructureerde regularisatie van lineaire en
multilineaire modellen verhoogt de interpretateerbaarheid
en verlaagt de trainingstijd en computationele complexiteit.
Dit draagt bij aan efficiëntere training en generalisatie, wat
de ontwikkeling van betrouwbaardere systemen toelaat, aangepast aan
verschillende contexten en gebruikersbehoeften.

>>>>

Naast onafhankelijkheid van oogbewegingen, verbetert dit werk ook algemene
ERP decodering door de structuur van lineaire en multilineaire technieken te verfijnen.
Deze methoden verhogen de nauwkeurigheid in verschillende BCI-toepassingen,
vooral wanneer kalibratiegegevens beperkt zijn.
Ge\-struc\-tu\-reerde regularisatie van lineaire en
multilineaire modellen verhoogt de interpretateerbaarheid
en verlaagt de trainingstijd en computationele complexiteit.
Dit draagt bij aan efficiëntere training en generalisatie, wat
de ontwikkeling van betrouwbaardere systemen toelaat, aangepast aan
verschillende contexten en gebruikersbehoeften.

### 2024-11-30 chapters/bci.tex
Layout chapter stbf-struct
Preface, ironed out overhangs

\fullpagefig[topleft]{%
  \includegraphics[width=\fullpagefigwidth]{figures/bci/bci_loop.pdf}
}{%
  \caption[The \acs{bci} loop.]{%
    The \ac{bci} loop.
    The user interacts with the \ac{bci} via a specific paradigm, in this case involving
    visual stimulation.
    Electrical neural signals are recorded, and neurophysiological features related to the
    paradigm are extracted.
    Using machine learning, a decision can then be made based on these features, which can
    be presented back to the user.
    In a closed-loop \ac{bci}, this feedback allows the user to adapt to the \ac{bci}.
  }
  \label{fig:bci/loop}
}

@@

\fullpagefig[topleft]{%
  \includegraphics[width=\fullpagefigwidth]{figures/bci/recording_modalities.pdf}
}{%
  \caption[Various electrophysiology recording technologies.]{%
	  Various electrophysiology recording technologies and their respective position in a
	  cross-section of the skull.
	  There is a trade-off between invasiveness and \\ spatial resolution.
  }
  \label{fig:bci/recording}
}

@@

In contrast to paradigms that require the user's active participation, for
instance through redirecting attention or initiating imagined actions,
paradigms free of this requirement are classified as \emph{passive \ac{bci}}.
From the user's point of view, this concept is probably the most attractive, since their
cognitive state would be inferred from just their brain activity.
However, without active user participation, it can be hard to establish reliable
communication or control.
Therefore, passive \ac{bci} are currently used for control tasks that require
input, like emotional affect~\cite{Torres2020,Libert2019, Muehl2014},
workload~\cite{Zanetti2021} or fatigue~\cite{Alimardani2020} detection,
or, more dependent on stimulation, error detection~\cite{SiMohammed2020}.
Aside from control, clinical applications are found in the diagnosis of a multitude of neurological
and psychological disorders through various forms of neuroimaging or monitoring,
in resting state or while performing a specific task.
Nevertheless, active participation facilitates training data collection for supervised
machine learning, since the conditions can be more easily controlled by the user or the
stimulation.

@@

The \ac{bci} paradigms mentioned above can also be combined to gain performance.
Straightforward examples are activating or deactivating \ac{ssvep} stimulation with a
motor command~\cite{Neeling2019}, or combining multiple visual paradigms by stimulating
along multiple dimensions at once.
\textcite{Han2023} recently used this strategy, combining frequency and phase coding in
\ac{ssvep} with the \ac{mvep} and oddball paradigms to develop an efficient \ac{bci}
where one of 200 targets can be accurately selected with only 800 ms of stimulation.

@@

Band-pass filtering is then usually applied to isolate the frequency ranges of interest
by removing frequencies that are too high or too low to contain relevant neural signals.
Band-pass filters eliminate slow drifts and high-frequency noise, such as muscle
artifacts or environmental electrical interference, improving the \ac{snr}.
Additionally, notch filtering can be used to remove strong power line interference\footnote{%
Typically at 50 Hz in Europe.}.
This interference can heavily contaminate \ac{eeg} recordings and even leak through when
outside the passband of a previous filter.
Notch filtering efficiently attenuates this narrowband noise without affecting the
surrounding frequencies.

Finally, to address artifacts like eye blinks and saccades that can significantly
distort the \ac{eeg} signal, \ac{ica} is performed~\cite{Delorme2007}.
\ac{ica} is a blind source separation technique that decomposes the \ac{eeg} data into
statistically independent components.
Components corresponding to eye movements, blinks, and other bodily or environmental
artifacts can be identified, either visually based on their characteristic topographies
and time courses, or statistically, and then removed.
This leads to cleaner data for subsequent processing, especially in the case of visual
\ac{bci} paradigms.
Since eye movement plays a role in the paradigm, it can strongly be present in the
signal and correlate with the task, forming a confounding factor for studying control
based solely on brain activity.

@@

This leads to a problem that is not often addressed:
a visual \ac{bci} is dependent on the gaze of the user, and hence on muscle control.
This is exactly what \acp{bci} for augmented and assistive technology try to avoid.
The \ac{bci} target population consists of individuals with severe physical
impairment due to  with various degrees
of paralysis, or even with \ac{lis}, the complete loss of muscle control with preserved
consciousness.
While \ac{bci}s are most attractive as a solution for individuals with \ac{sspi},
studies paradoxically often report poor performance in exactly this group.
This is caused by a multitude of factors, including diminished
electrophysiological responses and cognitive defects linked to the
pathophysiology, psychological factors such as the ``extinction of goal directed thinking''
phenomenon in complete impairment, and, importantly, sensory impairments such as vision and
oculomotor defects~\cite{Seguin2023}.
While visual \acp{bci} often rely on visual sensory functions, it is precisely this
group of patients that could benefit from them, that lacks adequate eye motor control.
Equally often, participants with gaze impairments were excluded from the study.

>>>>

\fullpagefig[caption.north]{%
  \includegraphics[width=\fullpagefigwidth]{figures/bci/bci_loop.pdf}
}{%
  \caption[The \acs{bci} loop.]{%
    The \ac{bci} loop.
    The user interacts with the \ac{bci} via a specific paradigm, in this case involving
    visual stimulation.
    Electrical neural signals are recorded, and neurophysiological features related to the
    paradigm are extracted.
    Using machine learning, a decision can then be made based on these features, which can
    be presented back to the user.
    In a closed-loop \ac{bci}, this feedback allows the user to adapt to the \ac{bci}.
  }
  \label{fig:bci/loop}
}

@@

\fullpagefig[caption.north]{%
  \includegraphics[width=\fullpagefigwidth]{figures/bci/recording_modalities.pdf}
}{%
  \caption[Various electrophysiology recording technologies.]{%
	  Various electrophysiology recording technologies and their respective position in a
	  cross-section of the skull.
	  There is a trade-off between invasiveness and spatial resolution.
  }
  \label{fig:bci/recording}
}

@@

In contrast to paradigms that require the user's active participation, for
instance through redirecting attention or initiating imagined actions,
pa\-ra\-digms free of this requirement are classified as \emph{passive \ac{bci}}.
From the user's point of view, this concept is probably the most attractive, since their
cognitive state would be inferred from just their brain activity.
However, without active user participation, it can be hard to establish reliable
communication or control.
Therefore, passive \ac{bci} are currently used for control tasks that require
input, like emotional affect~\cite{Torres2020,Libert2019, Muehl2014},
workload~\cite{Zanetti2021} or fatigue~\cite{Alimardani2020} detection,
or, more dependent on stimulation, error detection~\cite{SiMohammed2020}.
Aside from control, clinical applications are found in the diagnosis of a multitude of neurological
and psychological disorders through various forms of neuroimaging or monitoring,
in resting state or while performing a specific task.
Nevertheless, active participation facilitates training data collection for supervised
machine learning, since the conditions can be more easily controlled by the user or the
stimulation.

@@

The \ac{bci} paradigms mentioned above can also be combined to gain performance.
Straightforward examples are activating or deactivating \ac{ssvep} stimulation with a
motor command~\cite{Neeling2019}, or combining multiple visual pa\-ra\-digms by stimulating
along multiple dimensions at once.
\textcite{Han2023} recently used this strategy, combining frequency and phase coding in
\ac{ssvep} with the \ac{mvep} and oddball paradigms to develop an efficient \ac{bci}
where one of 200 targets can be accurately selected with only 800 ms of stimulation.

@@

Band-pass filtering is then usually applied to isolate the frequency ranges of interest
by removing frequencies that are too high or too low to contain relevant neural signals.
Band-pass filters eliminate slow drifts and high-fre\-quen\-cy noise, such as muscle
artifacts or environmental electrical interference, improving the \ac{snr}.
Additionally, notch filtering can be used to remove strong power line interference\footnote{%
Typically at 50 Hz in Europe.}.
This interference can heavily contaminate \ac{eeg} recordings and even leak through when
outside the passband of a previous filter.
Notch filtering efficiently attenuates this narrowband noise without affecting the
surrounding frequencies.

Finally, to address artifacts like eye blinks and saccades that can significantly
distort the \ac{eeg} signal, \ac{ica} is performed~\cite{Delorme2007}.
\ac{ica} is a blind source separation technique that decomposes the \ac{eeg} data into
statistically independent components.
Components corresponding to eye movements, eye blinks, and other bodily or environmental
artifacts can be identified, either visually based on their characteristic topographies
and time courses, or statistically, and then removed.
This leads to cleaner data for subsequent processing, especially in the case of visual
\ac{bci} paradigms.
Since eye movement plays a role in the paradigm, it can strongly be present in the
signal and correlate with the task, forming a confounding factor for studying control
based solely on brain activity.

@@

This leads to a problem that is not often addressed:
a visual \ac{bci} is dependent on the gaze of the user, and hence on muscle control.
This is exactly what \acp{bci} for augmented and assistive technology try to avoid.
The \ac{bci} target population consists of individuals with severe physical
impairment due to  with various degrees
of paralysis, or even with \ac{lis}, the complete loss of muscle control with preserved
consciousness.
While \ac{bci}s are most attractive as a solution for individuals with \ac{sspi},
studies paradoxically often report poor performance in exactly this group.
This is caused by a multitude of factors, including diminished
electrophysiological responses and cognitive defects linked to the
pathophysiology, psychological factors such as the ``extinction of goal directed thinking''
phe\-no\-me\-non in complete impairment, and, importantly, sensory impairments such as vision and
oculomotor defects~\cite{Seguin2023}.
While visual \acp{bci} often rely on visual sensory functions, it is precisely this
group of patients that could benefit from them, that lacks adequate eye motor control.
Equally often, participants with gaze impairments were excluded from the study.
